---
title: "The Changing Face of America"
subtitle: "Exploratory Analysis of County-Level Demographic Shifts (2019-2024)"
author: "Jonatan Tobón González"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r libraries, eval=FALSE}
install.packages("tidyverse")
install.packages("here")
install.packages("scales")
```


```{r load-libraries}
library(tidyverse)
library(here)
library(scales)
```

# Data Acquisition

The five datasets used in this analysis come form the U.S. Census Bureau's American Community Survey (ACS) 1-Year Demographic Profile table DP05, covering county-level estimates for 2019, 2021, 2022, 2023, and 2024. The year 2020 is absent because the Census Bureau suspended data collection during the COVID-19 pandemic. 

Each file is read with `col_types = cols(.default = "c")` to force all columns to character type on load. This is intentional because the ACS file contain suppresion codes like `"N"`, `"(x)"`, and `"-"` in numeric columns that would be silently lost if R attempted type inference at read time. Those values are handled explictily in the cleaning phase.

```{r load-data}
df_2019 <- read_csv(
  here("data", "raw", "2019", "ACSDP1Y2019.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2021 <- read_csv(
  here("data", "raw", "2021", "ACSDP1Y2021.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2022 <- read_csv(
  here("data", "raw", "2022", "ACSDP1Y2022.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2023 <- read_csv(
  here("data", "raw", "2023", "ACSDP1Y2023.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2024 <- read_csv(
  here("data", "raw", "2024", "ACSDP1Y2024.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)
```

# Initial Exploration

Before making any changes to the raw files, we take a first look at the structure and contents of each dataset. This step helps us understand the shape of the data, confirm that all files loaded correctly, and identify anything unexpected before we start cleaning.

## Dimensions

```{r dimesions}
tibble(
  year = c(2019, 2021, 2022, 2023, 2024),
  rows = c(
    nrow(df_2019),
    nrow(df_2021),
    nrow(df_2022),
    nrow(df_2023),
    nrow(df_2024)
  ),
  cols = c(
    ncol(df_2019),
    ncol(df_2021),
    ncol(df_2022),
    ncol(df_2023),
    ncol(df_2024)
  )
)
```

The row count increases each year from 841 to 862, reflecting ACS 1-Year survey's population threshold requirement; only counties with 65,000 or more residents are included, and more counties cross that threshold over time. The column count grows more sharply, from 359 in 2019 to 435 in 2024, due to structural changes the Census Bureau made ot its race and ethnicity question following the 2020 Census. This means we cannot simply select the same column codes across all five years and will need to map each variable to its correct code per year.

## Structure

```{r structure}
glimpse(df_2019)
```

## The Label Row

```{r label-row}
df_2019[1, 1:5]
```
Every DP05 export includes a second row containing verbose column description such as `"Estimate!!SEX AND AGE!!Total Population"` rather than actual data values. This row must be removed before any transformation or combination step, otherwise it would be treated as a data observation. 

# Cleaning & Transformation

## Preserving the Raw Data

Before making any changes we create a working copy of each dataframe. This ensures the original objects loaded from disk remain untouched and can be referenced at any point if needed. 

```{r preserve-originals}
raw_2019 <- df_2019
raw_2021 <- df_2021
raw_2022 <- df_2022
raw_2023 <- df_2023
raw_2024 <- df_2024
```

## Dropping the Label Row and Adding the Survey Year

The label row is removed from each dataframe using `slice(-1)`. Immediately after, we add a `survey_year` column and position it as the first column using `relocate()`. Both steps are chained together since there is no reason to separate them. 

```{r drop-label-add-year}
raw_2019 <- raw_2019 %>%
  slice(-1) %>%
  mutate(survey_year = 2019) %>%
  relocate(survey_year)
raw_2021 <- raw_2021 %>%
  slice(-1) %>%
  mutate(survey_year = 2021) %>%
  relocate(survey_year)
raw_2022 <- raw_2022 %>%
  slice(-1) %>%
  mutate(survey_year = 2022) %>%
  relocate(survey_year)
raw_2023 <- raw_2023 %>%
  slice(-1) %>%
  mutate(survey_year = 2023) %>%
  relocate(survey_year)
raw_2024 <- raw_2024 %>%
  slice(-1) %>%
  mutate(survey_year = 2024) %>%
  relocate(survey_year)
```

## Verifying the Label Row is Gone

```{r verify-label-row}
raw_2019[1, 1:5]
```

## Column selection and renaming

The 21 variables retained for this analysis must be selected from each year's file individually because Census Bureau variables codes shifted between 2019 and later years following updates to the race and ethnicity question. Where a code differs from the 2019 baseline, the shift is documented inline as a comment. 

```{r select-rename-2019}
raw_2019 <- raw_2019 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0071E,
    not_hispanic_latino_white_alone = DP05_0077E,
    not_hispanic_latino_black_alone = DP05_0078E,
    total_housing_units = DP05_0086E,
  )
```

```{r select-rename-2021}
raw_2021 <- raw_2021 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0071E,
    not_hispanic_latino_white_alone = DP05_0077E,
    not_hispanic_latino_black_alone = DP05_0078E,
    total_housing_units = DP05_0086E,
  )
```

```{r select-rename-2022}
raw_2022 <- raw_2022 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0073E, #Shifted from `DP05_0071E`
    not_hispanic_latino_white_alone = DP05_0079E, # Shifted from `DP05_0077E`
    not_hispanic_latino_black_alone = DP05_0080E, #Shifted from `DP05_0078E`
    total_housing_units = DP05_0088E, #Shifted from `DP05_0086E`
  )
```

```{r select-rename-2023}
raw_2023 <- raw_2023 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0047E, #Shifted from `DP05_0044E`
    native_hawaiian_pacific_islander_alone = DP05_0055E, #Shifted from `DP05_0052E`
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0076E, #Shifted from `DP05_0071E` and then again from `DP05_0073E`
    not_hispanic_latino_white_alone = DP05_0082E, # Shifted from `DP05_0077E` and then again from `DP05_0079E`
    not_hispanic_latino_black_alone = DP05_0083E, #Shifted from `DP05_0078E` and then again from `DP05_0080E`
    total_housing_units = DP05_0091E, #Shifted from `DP05_0086E` and then again from `DP05_0088E`
  )
```

```{r select-rename-2024}
raw_2024 <- raw_2024 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0045E, #Shifted from `DP05_0038E`
    american_indian_alaska_native_alone = DP05_0053E, #Shifted from `DP05_0039E`
    asian_alone = DP05_0061E, #Shifted from `DP05_0044E` and then again from `DP05_0047E`
    native_hawaiian_pacific_islander_alone = DP05_0069E, #Shifted from `DP05_0052E` and then again from `DP05_0055E`
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0090E, #Shifted from `DP05_0071E` and then again from `DP05_0073E` and one more time from `DP05_0076E`
    not_hispanic_latino_white_alone = DP05_0096E, # Shifted from `DP05_0077E` and then again from `DP05_0079E` and more time from `DP05_0082E`
    not_hispanic_latino_black_alone = DP05_0097E, #Shifted from `DP05_0078E` and then again from `DP05_0080E` and more time from `DP05_0083E`
    total_housing_units = DP05_0105E, #Shifted from `DP05_0086E` and then again from `DP05_0088E` and more time from `DP05_0091E`
  )
```

## Verify Column Selection

We confirm that all five dataframes now share identical column names and the expected dimensions before combining them. 

```{r verify-columns}
identical(names(raw_2019), names(raw_2021)) &
  identical(names(raw_2021), names(raw_2022)) &
  identical(names(raw_2022), names(raw_2023)) &
  identical(names(raw_2023), names(raw_2024))
```

```{r verify-dimensions}
tibble(
  year = c(2019, 2021, 2022, 2023, 2024),
  rows = c(
    nrow(df_2019),
    nrow(df_2021),
    nrow(df_2022),
    nrow(df_2023),
    nrow(df_2024)
  ),
  cols = c(
    ncol(df_2019),
    ncol(df_2021),
    ncol(df_2022),
    ncol(df_2023),
    ncol(df_2024)
  )
)

```
## Combining into a Single Dataset

With column names aligned across all five years, we stack them into a single longitudinal dataframe using `bind_rows()`. 

```{r combine}
census_raw <- bind_rows(raw_2019, raw_2021, raw_2022, raw_2023, raw_2024)

glimpse(census_raw)

```
## Special Value Replacement and Type Conversion

The ACS file use several placeholder values in numeric columns to indicate suppressed or unavailable estimates. Based on the Census Bureau's official table notes, these include `"N"` (insufficient sample cases), `"(X)"` (not applicable), and `"-"` (estimate could not be computed), as well as empty strings and numeric placeholders like `-888888888` and `-666666666`. We replace all of these with `NA` before converting numeric columns from character to double.

The column `geo_id` and `name` are excluded from conversion since they are intentionally character. 

```{r type-conversion}
non_numeric_cols <- c("geo_id", "name")

cols_to_convert <- census_raw %>%
  select(-all_of(non_numeric_cols)) %>%
  select(where(is.character)) %>%
  names()

census_clean <- census_raw %>%
  mutate(across(
    all_of(cols_to_convert),
    ~ ifelse(. %in% c("N", "(X)", "-", "", "NA", "-888888888", "-666666666"), NA, .)
  )) %>%
  mutate(across(
    all_of(cols_to_convert),
    as.numeric
  ))
```

## Verify Type Conversion

```{r verify-type}
census_clean %>%
  select(-geo_id, -name) %>%
  summarise(across(everything(), class)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "type")
```

```{r verify-na-counts}
census_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  filter(na_count > 0) %>%
  arrange(desc(na_count))
```
The missingness pattern is concentrated in racial composition columns. The 76 missing values shared across all racial breakdown columns represent county-year observations where the Census Bureau suppressed the full racial profile, likely due to small sample sizes. The `not_hispanic_latino_white_alone` and `not_hispanic_latino_black_alone` columns carry additional suppression (180 each), suggesting the Hispanic/non-Hispanic ethnicity breakdown was withheld from a broader set of counties. Core variables including `total_population`, `median_age`, `sex_ratio`, and `total_housing_units` are fully complete across all 4,244 observations. 

## Geographic Parsing

The `name` column cotains values formatted as `"County Name, State Name"`. We split this into two separate columns and immediately normalize `county_name` by stripping geographic suffixes (County, Parish, Borough, Municipality, and Census Area) that vary across records but carry no analytical value. 

```{r geographic-parsing}
census_clean <- census_clean %>%
  separate(name, into = c("county_name", "state_name"), sep = ", ") %>%
  mutate(county_name = str_remove(county_name, " County| Parish| Borough| Municipality| Census Area"))
```

```{r verify-parsing}
census_clean %>%
  select(county_name, state_name) %>%
  distinct() %>%
  arrange(state_name) %>%
  head(15)
```

## Regional Classification

We assign each state to one of the four U.S. Census Bureau regions and nine divisions using `case_when()`. Puerto Rico is classified separately as a Territory since it falls outside of the standard regional framework.

```{r regional-classification}
census_clean <- census_clean %>%
  mutate(
    region = case_when(
      state_name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")
      ~ "Northeast",
      state_name %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "Midwest",
      state_name %in% c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "District of Columbia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      state_name %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington") ~ "West",
      state_name == "Puerto Rico" ~ "Territory",
      TRUE ~ NA_character_
    ),
    division = case_when(
      state_name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont") ~ "New England",
      state_name %in% c("New Jersey", "New York", "Pennsylvania") ~ "Middle Atlantic",
      state_name %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin") ~ "East North Central",
      state_name %in% c("Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "West North Central",
      state_name %in% c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "District of Columbia", "West Virginia") ~ "South Atlantic",
      state_name %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee") ~ "East South Central",
      state_name %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "West South Central",
      state_name %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming") ~ "Mountain",
      state_name %in% c("Alaska", "California", "Hawaii", "Oregon", "Washington") ~ "Pacific",
      state_name == "Puerto Rico" ~ "Territory",
      TRUE ~ NA_character_
    )
  )
```

```{r verify-regional}
census_clean %>%
  distinct(state_name, region, division) %>%
  arrange(region, division, state_name)
```

