---
title: "The Changing Face of America"
author: "Jonatan Tobón González"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    df_print: paged
subtitle: "Exploratory Analysis of County-Level Demographic Shifts (2019-2024)"
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r libraries, eval=FALSE}
install.packages("tidyverse")
install.packages("here")
install.packages("scales")
```


```{r load-libraries}
library(tidyverse)
library(here)
library(scales)
```

```{r plot-theme}
theme_acs <- theme_minimal()+
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption = element_text(color = "gray50"),
    panel.grid = element_blank()
  )

theme_set(theme_acs)
```

# Data Acquisition

The five datasets used in this analysis come form the U.S. Census Bureau's American Community Survey (ACS) 1-Year Demographic Profile table DP05, covering county-level estimates for 2019, 2021, 2022, 2023, and 2024. The year 2020 is absent because the Census Bureau suspended data collection during the COVID-19 pandemic. 

Each file is read with `col_types = cols(.default = "c")` to force all columns to character type on load. This is intentional because the ACS file contain suppresion codes like `"N"`, `"(x)"`, and `"-"` in numeric columns that would be silently lost if R attempted type inference at read time. Those values are handled explictily in the cleaning phase.

```{r load-data}
df_2019 <- read_csv(
  here("data", "raw", "2019", "ACSDP1Y2019.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2021 <- read_csv(
  here("data", "raw", "2021", "ACSDP1Y2021.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2022 <- read_csv(
  here("data", "raw", "2022", "ACSDP1Y2022.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2023 <- read_csv(
  here("data", "raw", "2023", "ACSDP1Y2023.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)

df_2024 <- read_csv(
  here("data", "raw", "2024", "ACSDP1Y2024.DP05-Data.CSV"),
  col_types = cols(.default = "c"),
  show_col_types = FALSE
)
```

# Initial Exploration

Before making any changes to the raw files, we take a first look at the structure and contents of each dataset. This step helps us understand the shape of the data, confirm that all files loaded correctly, and identify anything unexpected before we start cleaning.

## Dimensions

```{r dimesions}
tibble(
  year = c(2019, 2021, 2022, 2023, 2024),
  rows = c(
    nrow(df_2019),
    nrow(df_2021),
    nrow(df_2022),
    nrow(df_2023),
    nrow(df_2024)
  ),
  cols = c(
    ncol(df_2019),
    ncol(df_2021),
    ncol(df_2022),
    ncol(df_2023),
    ncol(df_2024)
  )
)
```

The row count increases each year from 841 to 862, reflecting ACS 1-Year survey's population threshold requirement; only counties with 65,000 or more residents are included, and more counties cross that threshold over time. The column count grows more sharply, from 359 in 2019 to 435 in 2024, due to structural changes the Census Bureau made ot its race and ethnicity question following the 2020 Census. This means we cannot simply select the same column codes across all five years and will need to map each variable to its correct code per year.

## Structure

```{r structure}
glimpse(df_2019)
```

## The Label Row

```{r label-row}
df_2019[1, 1:5]
```
Every DP05 export includes a second row containing verbose column description such as `"Estimate!!SEX AND AGE!!Total Population"` rather than actual data values. This row must be removed before any transformation or combination step, otherwise it would be treated as a data observation. 

# Cleaning & Transformation

## Preserving the Raw Data

Before making any changes we create a working copy of each dataframe. This ensures the original objects loaded from disk remain untouched and can be referenced at any point if needed. 

```{r preserve-originals}
raw_2019 <- df_2019
raw_2021 <- df_2021
raw_2022 <- df_2022
raw_2023 <- df_2023
raw_2024 <- df_2024
```

## Dropping the Label Row and Adding the Survey Year

The label row is removed from each dataframe using `slice(-1)`. Immediately after, we add a `survey_year` column and position it as the first column using `relocate()`. Both steps are chained together since there is no reason to separate them. 

```{r drop-label-add-year}
raw_2019 <- raw_2019 %>%
  slice(-1) %>%
  mutate(survey_year = 2019) %>%
  relocate(survey_year)
raw_2021 <- raw_2021 %>%
  slice(-1) %>%
  mutate(survey_year = 2021) %>%
  relocate(survey_year)
raw_2022 <- raw_2022 %>%
  slice(-1) %>%
  mutate(survey_year = 2022) %>%
  relocate(survey_year)
raw_2023 <- raw_2023 %>%
  slice(-1) %>%
  mutate(survey_year = 2023) %>%
  relocate(survey_year)
raw_2024 <- raw_2024 %>%
  slice(-1) %>%
  mutate(survey_year = 2024) %>%
  relocate(survey_year)
```

## Verifying the Label Row is Gone

```{r verify-label-row}
raw_2019[1, 1:5]
```

## Column selection and renaming

The 21 variables retained for this analysis must be selected from each year's file individually because Census Bureau variables codes shifted between 2019 and later years following updates to the race and ethnicity question. Where a code differs from the 2019 baseline, the shift is documented inline as a comment. 

```{r select-rename-2019}
raw_2019 <- raw_2019 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0071E,
    not_hispanic_latino_white_alone = DP05_0077E,
    not_hispanic_latino_black_alone = DP05_0078E,
    total_housing_units = DP05_0086E,
  )
```

```{r select-rename-2021}
raw_2021 <- raw_2021 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0071E,
    not_hispanic_latino_white_alone = DP05_0077E,
    not_hispanic_latino_black_alone = DP05_0078E,
    total_housing_units = DP05_0086E,
  )
```

```{r select-rename-2022}
raw_2022 <- raw_2022 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0044E,
    native_hawaiian_pacific_islander_alone = DP05_0052E,
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0073E, #Shifted from `DP05_0071E`
    not_hispanic_latino_white_alone = DP05_0079E, # Shifted from `DP05_0077E`
    not_hispanic_latino_black_alone = DP05_0080E, #Shifted from `DP05_0078E`
    total_housing_units = DP05_0088E, #Shifted from `DP05_0086E`
  )
```

```{r select-rename-2023}
raw_2023 <- raw_2023 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0038E,
    american_indian_alaska_native_alone = DP05_0039E,
    asian_alone = DP05_0047E, #Shifted from `DP05_0044E`
    native_hawaiian_pacific_islander_alone = DP05_0055E, #Shifted from `DP05_0052E`
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0076E, #Shifted from `DP05_0071E` and then again from `DP05_0073E`
    not_hispanic_latino_white_alone = DP05_0082E, # Shifted from `DP05_0077E` and then again from `DP05_0079E`
    not_hispanic_latino_black_alone = DP05_0083E, #Shifted from `DP05_0078E` and then again from `DP05_0080E`
    total_housing_units = DP05_0091E, #Shifted from `DP05_0086E` and then again from `DP05_0088E`
  )
```

```{r select-rename-2024}
raw_2024 <- raw_2024 %>%
  select(
    survey_year,
    geo_id = GEO_ID,
    name = NAME,
    total_population = DP05_0001E,
    total_male_population = DP05_0002E,
    total_female_population = DP05_0003E,
    sex_ratio = DP05_0004E,
    population_under_5 = DP05_0005E,
    population_under_18 = DP05_0019E,
    population_16_and_over = DP05_0020E,
    population_18_and_over = DP05_0021E,
    population_65_and_over = DP05_0024E,
    median_age = DP05_0018E,
    white_alone = DP05_0037E,
    black_alone = DP05_0045E, #Shifted from `DP05_0038E`
    american_indian_alaska_native_alone = DP05_0053E, #Shifted from `DP05_0039E`
    asian_alone = DP05_0061E, #Shifted from `DP05_0044E` and then again from `DP05_0047E`
    native_hawaiian_pacific_islander_alone = DP05_0069E, #Shifted from `DP05_0052E` and then again from `DP05_0055E`
    two_or_more_races = DP05_0035E,
    hispanic_latino = DP05_0090E, #Shifted from `DP05_0071E` and then again from `DP05_0073E` and one more time from `DP05_0076E`
    not_hispanic_latino_white_alone = DP05_0096E, # Shifted from `DP05_0077E` and then again from `DP05_0079E` and more time from `DP05_0082E`
    not_hispanic_latino_black_alone = DP05_0097E, #Shifted from `DP05_0078E` and then again from `DP05_0080E` and more time from `DP05_0083E`
    total_housing_units = DP05_0105E, #Shifted from `DP05_0086E` and then again from `DP05_0088E` and more time from `DP05_0091E`
  )
```

## Verify Column Selection

We confirm that all five dataframes now share identical column names and the expected dimensions before combining them. 

```{r verify-columns}
identical(names(raw_2019), names(raw_2021)) &
  identical(names(raw_2021), names(raw_2022)) &
  identical(names(raw_2022), names(raw_2023)) &
  identical(names(raw_2023), names(raw_2024))
```

```{r verify-dimensions}
tibble(
  year = c(2019, 2021, 2022, 2023, 2024),
  rows = c(
    nrow(df_2019),
    nrow(df_2021),
    nrow(df_2022),
    nrow(df_2023),
    nrow(df_2024)
  ),
  cols = c(
    ncol(df_2019),
    ncol(df_2021),
    ncol(df_2022),
    ncol(df_2023),
    ncol(df_2024)
  )
)

```
## Combining into a Single Dataset

With column names aligned across all five years, we stack them into a single longitudinal dataframe using `bind_rows()`. 

```{r combine}
census_raw <- bind_rows(raw_2019, raw_2021, raw_2022, raw_2023, raw_2024)

glimpse(census_raw)

```
## Special Value Replacement and Type Conversion

The ACS file use several placeholder values in numeric columns to indicate suppressed or unavailable estimates. Based on the Census Bureau's official table notes, these include `"N"` (insufficient sample cases), `"(X)"` (not applicable), and `"-"` (estimate could not be computed), as well as empty strings and numeric placeholders like `-888888888` and `-666666666`. We replace all of these with `NA` before converting numeric columns from character to double.

The column `geo_id` and `name` are excluded from conversion since they are intentionally character. 

```{r type-conversion}
non_numeric_cols <- c("geo_id", "name")

cols_to_convert <- census_raw %>%
  select(-all_of(non_numeric_cols)) %>%
  select(where(is.character)) %>%
  names()

census_clean <- census_raw %>%
  mutate(across(
    all_of(cols_to_convert),
    ~ ifelse(. %in% c("N", "(X)", "-", "", "NA", "-888888888", "-666666666"), NA, .)
  )) %>%
  mutate(across(
    all_of(cols_to_convert),
    as.numeric
  ))
```

## Verify Type Conversion

```{r verify-type}
census_clean %>%
  select(-geo_id, -name) %>%
  summarise(across(everything(), class)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "type")
```

```{r verify-na-counts}
census_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  filter(na_count > 0) %>%
  arrange(desc(na_count))
```
The missingness pattern is concentrated in racial composition columns. The 76 missing values shared across all racial breakdown columns represent county-year observations where the Census Bureau suppressed the full racial profile, likely due to small sample sizes. The `not_hispanic_latino_white_alone` and `not_hispanic_latino_black_alone` columns carry additional suppression (180 each), suggesting the Hispanic/non-Hispanic ethnicity breakdown was withheld from a broader set of counties. Core variables including `total_population`, `median_age`, `sex_ratio`, and `total_housing_units` are fully complete across all 4,244 observations. 

## Geographic Parsing

The `name` column cotains values formatted as `"County Name, State Name"`. We split this into two separate columns and immediately normalize `county_name` by stripping geographic suffixes (County, Parish, Borough, Municipality, and Census Area) that vary across records but carry no analytical value. 

```{r geographic-parsing}
census_clean <- census_clean %>%
  separate(name, into = c("county_name", "state_name"), sep = ", ") %>%
  mutate(county_name = str_remove(county_name, " County| Parish| Borough| Municipality| Census Area"))
```

```{r verify-parsing}
census_clean %>%
  select(county_name, state_name) %>%
  distinct() %>%
  arrange(state_name) %>%
  head(15)
```

## Regional Classification

We assign each state to one of the four U.S. Census Bureau regions and nine divisions using `case_when()`. Puerto Rico is classified separately as a Territory since it falls outside of the standard regional framework.

```{r regional-classification}
census_clean <- census_clean %>%
  mutate(
    region = case_when(
      state_name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")
      ~ "Northeast",
      state_name %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "Midwest",
      state_name %in% c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "District of Columbia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      state_name %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington") ~ "West",
      state_name == "Puerto Rico" ~ "Territory",
      TRUE ~ NA_character_
    ),
    division = case_when(
      state_name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont") ~ "New England",
      state_name %in% c("New Jersey", "New York", "Pennsylvania") ~ "Middle Atlantic",
      state_name %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin") ~ "East North Central",
      state_name %in% c("Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota") ~ "West North Central",
      state_name %in% c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "District of Columbia", "West Virginia") ~ "South Atlantic",
      state_name %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee") ~ "East South Central",
      state_name %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "West South Central",
      state_name %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming") ~ "Mountain",
      state_name %in% c("Alaska", "California", "Hawaii", "Oregon", "Washington") ~ "Pacific",
      state_name == "Puerto Rico" ~ "Territory",
      TRUE ~ NA_character_
    )
  )
```

```{r verify-regional}
census_clean %>%
  distinct(state_name, region, division) %>%
  arrange(region, division, state_name)
```

# Exploratory Data Analysis

Before calculating any derived variables, we explore the combined dataset to understand its structure, distributions, and patterns. Each exploration informs the decisions we make in the next phase.

## Population Distribution

County population in this dataset are highly skewed (a small number of very large counties coexist with a large number of small ones). We look at the overall distribution first, then break it down by region. 

```{r population-distribution}
census_clean %>%
  filter(survey_year == 2024) %>%
  ggplot(aes(x = total_population))+
  geom_histogram(bins = 50, fill = "steelblue", color = "white")+
  scale_x_continuous(labels = comma)+
  labs(
    title = "Distribution of County Population (2024)",
    subtitle = "Each bin represents a range of county population sizes",
    x = "Total Population",
    y = "Number of Counties",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
The raw distribution makes the skew immediately visible. The vast majority of counties in the dataset cluster below 500,000 residents, while a handful of very large counties (Los Angeles, Cook, Harris) stretch the x-axis out to nearly 10 million. This kind of right skew is typical of county population data in the United States and has an important implication for analysis: any regional average that treats counties as equal units will be dominated by the largest ones.

```{r population-log}
census_clean %>%
  filter(survey_year == 2024) %>%
  ggplot(aes(x = total_population))+
  geom_histogram(bins = 50, fill = "steelblue", color = "white")+
  scale_x_log10(labels = comma)+
  labs(
    title = "Distribution of County Population on Log Scale (2024)",
    subtitle = "Log scale reveals the full range of county sizes more clearly",
    x = "Total Population (Log Scale)",
    y = "Number of Counties",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
On the log scale the distribution becomes much more informative. County populations are spread fairly continuously from roughly 65,000 (The ACS 1-Year threshold) up through the millions, with the bulk of counties falling between 100,000 and 500,000 residents. The log scale will be the appropriate choice whenever we plot population on an axis going forward. 

```{r population-summary}
census_clean %>%
  filter(survey_year == 2024) %>%
  summarise(
    min = min(total_population, na.rm = TRUE),
    q25 = quantile(total_population, 0.25, na.rm = TRUE),
    median = median(total_population, na.rm = TRUE),
    mean = mean(total_population, na.rm = TRUE),
    q75 = quantile(total_population, 0.75, na.rm = TRUE),
    max = max(total_population, na.rm = TRUE)
  )

```
The summary confirms the skew numerically. The median county population in 2024 is 163,839 while the mean is 340,954 (nearly double). This gap between median and mean is the fingerprint of a right-skewed distribution and tells us that a small number of very large counties are pulling the average upward significantly. 

## Median Age Distribution

```{r age-distribution}
census_clean %>%
  filter(survey_year == 2024) %>%
  ggplot(aes(x = median_age))+
  geom_histogram(bins = 40, fill = "steelblue", color = "white")+
  labs(
    title = "Distribution of County Median Age (2024)",
    subtitle = "Each bin represents a range of median ages across counties",
    x = "Median Age",
    y = "Number of Counties",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
The overall distribution of county median age is approximately normal and centered around 40 to 42 years, with a right tail extending toward counties with median age above 55. These high-age outliers likely represent retirement communities or rural counties experiencing sustained outmigration of younger residents.

```{r age-by-region}
census_clean %>%
  filter(survey_year == 2024) %>%
  ggplot(aes(x = median_age, fill = region))+
  geom_density(alpha = 0.4)+
  labs(
    title = "Median Age Distribution by Region (2024)",
    subtitle = "Density curves show how age distributions differ across Census regions",
    x = "Median Age",
    y = "Density",
    fill = "Region",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
The regional density plot reveals a pattern that raw averages would obscure. Puerto Rico stands apart with a narrow, sharply peaked distribution centered around 47 to 48 years (its counties cluster tightly around an older median age than any continental region and show far less internal variation). The continental regions overlap considerably in the 37 to 44 range, though the Northeast skews slightly older and the South and West show broader, flatter distributions suggesting more internal diversity in age structure.

```{r age-summary-by-region}
census_clean %>%
  filter(survey_year == 2024) %>%
  group_by(region) %>%
  summarise(
    min = min(median_age, na.rm = TRUE),
    q25 = quantile(median_age, 0.25, na.rm = TRUE),
    median = median(median_age, na.rm = TRUE),
    mean = mean(median_age, na.rm = TRUE),
    q75 = quantile(median_age, 0.75, na.rm = TRUE),
    max = max(median_age, na.rm = TRUE)
  ) %>%
  arrange(desc(median))
```
The summary confirms Puerto Rico as the olderst region by median (44.9) despite being a single territory rather than a collection of diverse states. Among continental regions, the Northeast leads at 41.9, while the South and West both sits at 39.0. The maximum of 68.0 in the South is a notable outlier worth identifying. 

```{r age-outlier}
census_clean %>%
  filter(survey_year == 2024, region == "South") %>%
  arrange(desc(median_age)) %>%
  select(county_name, state_name, median_age, total_population) %>%
  head(5)
```
Sumter County, Florida emerges as the clear outlier at a median age of 68.0 with a total population of 154,693. This is not a data quality issue, Sumter County is home to The Villages, one of the largest planned retirement communities in the United States, and its age structure reflects that reality. This county will appear repeatedly as an outlier in age-related analyses throughout this notebook and should be interpreted in that context rather than treated as an anomaly to be removed. 

## Sex Ratio Distributions

Sex ration is expressed as the number of males per 100 females. A perfectly balanced county would show a value of 100. We first look at the overall distribution, then identify where the extremes are concentrated. 

```{r sex-ratio-distributions}
census_clean %>%
  filter(survey_year == 2024) %>%
  ggplot(aes(x = sex_ratio))+
  geom_histogram(bins = 40, fill = "steelblue", color = "white")+
  geom_vline(xintercept = 100, linetype = "dashed", color = "gray40")+
  labs(
    title = "Distribution of County Sex Ratios (2024)",
    subtitle = "Dashed line marks a balanced ratio of 100 males per 100 females",
    x = "Sex Ratio (Males per 100 Females",
    y = "Number of Counties",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
The distribution of county sex ratios is approximately normal and centered just below 100, meaning the typical county has slightly more females than males. The median of 97.8 and mean of 97.99 are nearly identical, confirming the distribution is symmetric for the majority of counties. The dashed line at 100 falls near the peak but slightly to the right, making the female-heavy skew of most counties visible.

```{r sex-ratio-summary}
census_clean %>%
  filter(survey_year == 2024) %>%
  summarise(
    min = min(sex_ratio, na.rm = TRUE),
    q25 = quantile(sex_ratio, 0.25, na.rm = TRUE),
    median = median(sex_ratio, na.rm = TRUE),
    mean = mean(sex_ratio, na.rm = TRUE),
    q75 = quantile(sex_ratio, 0.75, na.rm = TRUE),
    max = max(sex_ratio, na.rm = TRUE)
  )
```

```{r sex-ratio-outliers}
census_clean %>%
  filter(survey_year == 2024) %>%
  mutate(
    q1 = quantile(sex_ratio, 0.25, na.rm = TRUE),
    q3 = quantile(sex_ratio, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    lower = q1 - 1.5 * iqr,
    upper = q3 + 1.5 * iqr
  ) %>%
  filter(sex_ratio < lower | sex_ratio > upper) %>%
  select(county_name, state_name, region, sex_ratio, total_population) %>%
  arrange(desc(sex_ratio))
```
The 34 outlier counties identified by the 1.5 * IQR rule fall into two distinct and interpretable groups. The high-ratio outliers (led by Walker County, Texas (134.3), Fairbanks North Start Borough, Alaska (126.5), Kings County, California (124.2), and Onslow County, North Carolina (121.8)) are almost entirely explained by institutional population. Walker and Kings counties host large state and federal prisons, Fairbanks North Start and Onslow are home to major military installations. These are not demographic anomalies, they are accurate reflections of the population living in those counties at the time of the survey. 

The low-ratio outliers tell a different story. Three Puerto Rico municipios appear at the bottom of the distribution (82.6-85.0), consistent with the broader pattern of male outmigration from the island that has accelerated since Hurricane Maria in 2017. Baltimore City (86.4) and several majority-Black counties in the South follow, a patter that research has consitently linked to higher rates of male incareceration and premature mortality in those communities. Both the high and low extremes underscore that sex ratio outliers are rarely random, they reflect structural conditions worth investigating further.

## Missing Data by Region and Year

Earlier we identified that missingness is concentrated in racial composition columns. Here we explore whether that suppression is randomly distributed across regions and years or whether it follows a pattern.

```{r missing-by-year}
census_clean %>%
  group_by(survey_year) %>%
  summarise(across(
    c(white_alone, black_alone, asian_alone, native_hawaiian_pacific_islander_alone, american_indian_alaska_native_alone, two_or_more_races), ~ sum(is.na(.))
  )) %>%
  pivot_longer(-survey_year, names_to = "column", values_to = "na_count") %>%
  ggplot(aes(x = survey_year, y = na_count, color = column))+
  geom_line(linewidth = 1)+
  geom_point(size = 2)+
  scale_x_continuous(breaks = c(2019, 2021, 2022, 2023, 2024))+
  labs(
    title = "Missing Values in Racial Composition Columns by Year",
    subtitle = "Tracking Census suppression patterns across five survey years",
    x = "Survey Year",
    y = "Number of Missing Values",
    color = "Column",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
Two patterns emerge clearly from this exploration. First, suppression is heavily concentrated in 2019. The by-year chart shows roughly 60 missing values in `white_alone` and `two_or_more_races` in 2019, dropping sharply to single digits in every subsequent year. This is directly linked to the Census Bureau's redesign of the race and ethnicity question following the 2020 Census, the updated question captured racial identity more completely, reducing the number of records where racial composition could not be reported.

```{r missing-by-region}
census_clean %>%
  group_by(region) %>%
  summarise(across(
    c(white_alone, black_alone, asian_alone, native_hawaiian_pacific_islander_alone, american_indian_alaska_native_alone, two_or_more_races), ~ sum(is.na(.))
  )) %>%
  pivot_longer(-region, names_to = "column", values_to = "na_count") %>%
  ggplot(aes(x = region, y = na_count, fill = column))+
  geom_col(position = "dodge")+
  labs(
    title = "Missing Values in Racial Composition Columns by Region",
    subtitle = "Suppression patterns across Census regions",
    x = "Region",
    y = "Number of Missing Values",
    fill = "Column",
    caption = "Source: ACS DP05 1-Year Estimates"
  )
```
Second, the South accounts for the largest share of suppression by region, with approximately 48 missing values across racial columns, roughly half the total. This is proportional rather than structural: the South simply contains more counties than any other region (355 to 368 depending on the year), so it naturally accumulates more instances of suppression even when its suppression rate is lower than other regions. 

```{r missing-count-tables}
census_clean %>%
  group_by(region, survey_year) %>%
  summarise(
    missing_any_race = sum(is.na(white_alone) | is.na(black_alone) | is.na(asian_alone) | is.na(two_or_more_races)),
    total_counties = n(),
    pct_missing = round((missing_any_race / total_counties) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_missing))
```
The percent table confirms that suppression rates are low across the board. The highest rate observed is 18.2% for the Territory in 2019, which represents just 2 of 11 Puerto Rico municipios, a small absolute count that looks large only because the Territory has so few counties. Among continental regions, the South in 2019 at 9.9% is the only case approaching a meaningful suppression rate, and it drops to under 1% in every subsequent year. Given this pattern, missingness in this dataset is not randomly distributed, it is concentrated in 2019 and disproportionately affects the South, but its overall scale is small enough that it will not materially affect regional or longitudinal analysis. 
